---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# txendcutr

<!-- badges: start -->
[![R build status](https://github.com/guillermo1996/txendcutr/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/guillermo1996/txendcutr/actions)
[![codecov](https://codecov.io/gh/guillermo1996/txendcutr/branch/txendcutr-release/graph/badge.svg)](https://codecov.io/gh/guillermo1996/txendcutr)
<!-- badges: end -->

## Overview

Various mRNA sequencing library preparation methods generate sequencing reads
from the transcript ends. Quantification of isoform usage can be improved by
using truncated versions of transcriptome annotations when assigning such reads
to isoforms. The `txendcutr` package implements some convenience methods for
readily generating such truncated annotations from either their 5' or 3'
transcript ends and their corresponding sequences.

`txendcutr` is a standalone fork of
[`txcutr`](https://github.com/mfansler/txcutr), maintained independently of
the upstream Bioconductor package. See [Differences from
`txcutr`](#differences-from-txcutr) below for what has changed.

## Installation instructions

`txendcutr` is not published on Bioconductor or CRAN. Install it directly
from GitHub. Get the latest stable `R` release from
[CRAN](http://cran.r-project.org/), then install `txendcutr` with:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

## BiocManager::install() also installs from GitHub, and correctly
## resolves this package's Bioconductor dependencies
BiocManager::install("guillermo1996/txendcutr")
```

or, equivalently, with `remotes`:

```{r 'install_remotes', eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}

remotes::install_github("guillermo1996/txendcutr")
```

## Example

A typical workflow for `txendcutr` involves

 - loading an existing annotation as a `TxDb` object
 - truncating the annotation from the 3' or 5' end
 - exporting the truncated annotation (GTF)
 - exporting supporting files (FASTA, merge TSV)

```{r eval=FALSE}
library(rtracklayer)
library(txendcutr)
library(BSgenome.Hsapiens.UCSC.hg38)

## load human genome
hg38 <- BSgenome.Hsapiens.UCSC.hg38

## load human GENCODE annotation
txdb <- makeTxDbFromGFF("gencode.v38.annotaton.gtf.gz", organism="Homo sapiens")

## truncate to maximum of 500 nts from the 3' end
txdb_w500 <- truncate3primeTxome(txdb, maxTxLength=500)

## ...or from the 5' end
txdb_5p_w500 <- truncate5primeTxome(txdb, maxTxLength=500)

## export annotation
exportGTF(txdb_w500, file="gencode.v38.txendcutr_w500.gtf.gz")

## export FASTA
exportFASTA(txdb_w500, genome=hg38, file="gencode.v38.txendcutr_w500.fa.gz")

## export merge-table
exportMergeTable(txdb_w500, minDistance=200,
                 file="gencode.v38.txendcutr_w500.merge.tsv.gz")
```

## Differences from `txcutr`

`txendcutr` began as a fork of
[`mfansler/txcutr`](https://github.com/mfansler/txcutr) and has since
diverged as a standalone package. Key changes:

* **Native 5' truncation.** Upstream `txcutr` only truncates from the 3'
  end. `txendcutr` generalizes `truncateTxome()` with a `txEnd` argument
  (`"3prime"` or `"5prime"`), and adds `truncate3primeTxome()` /
  `truncate5primeTxome()` convenience wrappers (each with a `quiet` option
  to suppress progress messages).
* **Overlap export.** `truncateTxome()` (and both wrappers) accept an
  `overlapFile` argument to export a TSV of transcript pairs that were
  collapsed as duplicates during truncation, for auditing and debugging.
* **Faster truncation pipeline.** The internal clipping step no longer
  schedules one `BiocParallel` task per transcript; transcripts are now
  batched across workers and processed with a vectorized `dplyr` pipeline,
  which is substantially faster on large transcriptomes.

## Citation

`txendcutr` is a derivative work, built directly on the methodology and
original implementation of [`txcutr`](https://github.com/mfansler/txcutr)
by Mervin Fansler. **If you use `txendcutr`, please cite both packages.**

Below is the citation output from using `citation('txendcutr')` in R. Please
run this yourself to check for any updates on how to cite __txendcutr__.

```{r 'citation', eval = requireNamespace('txendcutr')}
print(citation('txendcutr'), bibtex = TRUE)
```

Please also cite the original `txcutr` package, without which this project
would not exist:

```{r 'citation_txcutr', echo = FALSE}
txcutr_citation <- utils::bibentry(
    bibtype = "Manual",
    title   = "txcutr: Transcriptome CUTteR",
    author  = "Mervin Fansler",
    year    = "2025",
    note    = "R package version 1.15.2",
    url     = "https://github.com/mfansler/txcutr"
)
print(txcutr_citation)
cat("\nA BibTeX entry for LaTeX users is\n\n")
print(utils::toBibtex(txcutr_citation))
```

Note that `txendcutr` was only made possible thanks to Mervin
Fansler's original work on `txcutr`, as well as many other R and
bioinformatics software authors, which are cited either in the vignettes
and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `txendcutr` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub Actions](https://github.com/guillermo1996/txendcutr/actions), running on [Bioconductor's Docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`, adapted from the workflow originally generated by `r BiocStyle::Biocpkg('biocthis')` for upstream `txcutr`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh/guillermo1996/txendcutr) and `r BiocStyle::CRANpkg('covr')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.
* Unit testing is powered by `r BiocStyle::CRANpkg('testthat')` (3rd edition).

This package originates from Mervin Fansler's `txcutr`; see the [upstream repository](https://github.com/mfansler/txcutr) for its own, separately maintained development infrastructure (including a `pkgdown` documentation site).

